Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-04-01 article-models-main.tex
added lots of references

In this part we develop type theory from the ground up. We start with a type
theory without any of the basic constructors. This is the theory of contexts
families and terms which has the basic operations of extension, weakening,
substitution and identity terms. Type theory before type constructors has not
been studied very much. Dependent product types or even universes tend to make
an early appearance in just about any presentation of type theory. While one
could argue that in type theories like the one in \cite{Dybjer1996} the type
constructors could be left out and a study of the remnants could \emph{in
principle} be carried out, this is not the usual practice. In the type theory
of \cite{TheBook}, which seems to have won the race of introducing universes
as early as possible hands down, this seems entirely unfeasible.

Nevertheless, type theory without type constructors has received some attention
recently. In \cite{Garner2014}, Garner describes the combinatorial structure
of the type operations of the weakening, substitution and projection monads
(their projections are our identity terms) and suggests lots of further research
that can be done on type theory without constructors. Also, Joyal has been
lecturing about tribes lately, which are a categorical semantics for type theory
without type constructors. Although his slides have been going around on
mailing lists, there is as of yet no easily accessible written accound on tribes.

>>>>

In this part we develop type theory from the ground up. We start with a type
theory without any of the basic constructors. This is the theory of contexts
families and terms which has the basic operations of extension, weakening,
substitution and identity terms. Type theory before type constructors has not
been studied very much. Dependent product types or even universes tend to make
an early appearance in just about any presentation of type theory. While one
could argue that in type theories like the one in \cite{Dybjer1996} the type
constructors could be left out and a study of the remnants could \emph{in
principle} be carried out, this is not the usual practice. A noteable exception
is the unpublished work \cite{Awodey2013} on natural models of type theory,
which makes the connection between categories with families and representable
transformations of presheaves. In the type theory
of \cite{TheBook}, which seems to have won the race of introducing universes
as early as possible hands down, this seems entirely unfeasible. Also, the 
proof-assistant \Coq\ {\color{red}(and \Agda too?)} has universes and dependent product and pair types 
built-in, making it impossible to study type theory without type constructors in 
that environment.

Nevertheless, type theory without type constructors has received some attention
contributors to the Univalent Foundations Program recently. We name two
further investigations on this topic, other than the mentioned work by Awodey.
In \cite{Garner2014}, Garner describes the combinatorial structure
of the type operations of the weakening, substitution and projection monads
(their projections are our identity terms) and suggests lots of further research
that can be done on type theory without constructors. Also, Joyal has been
lecturing about tribes lately, which are a categorical semantics for type theory
without type constructors. Although his slides have been going around on
mailing lists, there is as of yet no easily accessible written accound on tribes.

The theory that is layed out in this part is built up modularly. As announced,
we begin by presenting type theory without type constructors. This type theory
can be extended by adding type constructors. (Probably) There are three flavors
of type constructors. Dependent function types and universes are each of their
own flavor; the inductive types form the third flavor.

However the type theory can also be extended with local versions of extenison,
weakening, substitution and identity terms. This comprises a type theory of
(internal) models of the theory of contexts families and terms presented in
\autoref{tt} and it can be further extended to also model any combination of
type constructors.

### 2014-04-01 article-models-tt.tex
added intro to part II
added lots of references

\section{The theory of contexts, families and terms}\label{tt}
In this section we give a description of dependent type theory before type
constructors with explicit operations of extension, weakening and substitution. We will
formulate the type theory in such a way that contexts aren't defined
to be lists of variable
declarations. We rather take a syntactic approach in our treatment of contexts,
axiomatizing only what you can do with them and allowing for the freedom to
interpret the axiom in any desired way. 
The main reason is that we don't see variable declarations in the internal
models either. This way we also set out to a more algebraic approach of type
theory and higher category theory. Thirdly, we will not have to be burdened with
superficial comments about variables being bounded or not, or fresh or free or 
not occuring at all.

The theory we will develop here is a theory of contexts, families of contexts
and terms thereof. Families of contexts could have been named dependent contexts
or telescopes.

@@

>>>>

\section{The theory of contexts, families and terms}\label{tt}
In this section we give a description of dependent type theory before type
constructors. Apart from contexts, families and terms -- which provide for the
core of the language of dependent logic -- the basic ingredients
of this theory will be the operations of extension, weakening and substitution
and the identity terms. The resulting theory can be seen as the structure of
dependent type theory. We note up front that what we call families over contexts
here, could have been named dependent contexts or telescopes, see
\cite{deBruijn1991,hofmann1995extensional}.

We will
formulate the type theory in such a way that contexts aren't defined
to be lists of variable declarations. The variable-free (or name-free) approach 
we take here is rather different than those appearing in 
\cite{hofmann1995extensional,TheBook} but it has appeared in the work of Coquand
and in \cite{Dybjer1996}.
The main reason we don't let variable declarations in is that we don't see them 
in the internal models either. This way we also set out to a more algebraic 
approach of type theory and higher category theory. Thirdly, we will not have to
be burdened with superficial comments about variables being bounded or not, or 
fresh or free or not occuring at all.

@@

\begin{itemize}
\item Extensions of inductive morphisms are inductive
\item Weakenings of inductive morphisms are inductive
\item Substitutions of inductive morphisms are inductive
\end{itemize}
